Emergent dual scaling of riverine biodiversity

### 2021-03-26 manuscript_pnas.Rmd
Update manuscript_pnas.Rmd

Here, we propose a unified framework of ecosystem size and complexity in scaling biodiversity patterns in rivers - a prime example of complex branching ecosystems. Individual streams and rivers flow through different landscapes with distinct geological and climatic backgrounds, serving as a spatial unit of unique in-stream environments [@brownDendriticNetworkStructure2010; @petersonModellingDendriticEcological2013; @mooreEmergentStabilityLarge2015; @teruiMetapopulationStabilityBranching2018; @isaakSlowClimateVelocities2016; @mcguireNetworkAnalysisReveals2014; @abbottUnexpectedSpatialStability2018]. The recurrent merging of diverse tributaries ultimately forms a fluvial network with fractal branching patterns [@rodriguez-iturbeFractalRiverBasins2001]. As such, the complexity of branching structure, which we define here as the probability of branching per unit river distance [@teruiMetapopulationStabilityBranching2018; @yeakelSynchronisationStabilityRiver2014], may represent a macro-scale control of the ecosystem's habitat heterogeneity (habitat diversity per unit area) [@grantLivingBranchesPopulation2007; @mooreEmergentStabilityLarge2015; @teruiMetapopulationStabilityBranching2018]. Meanwhile, ecosystem size (watershed area) should determine the metacommunity size and total habitat diversity (area x heterogeneity), two factors that regulate biodiversity at the metacommunity level [@hubbellUnifiedNeutralTheory2001; @kallimanisHowDoesHabitat2008]. Hence, we hypothesize that riverine biodiversity obeys power laws along the two orthogonal dimensions of branching networks.

The present study combines theory and analysis of extensive community data from two different regions of the globe to show that multiple ecological pathways converge to the emergence of dual diversity scaling. Specifically, watershed-scale species richness ($\gamma$ diversity) increased linearly with ecosystem size $A$ and branching probability $P_b$ on logarithmic axes as $log_{10}~\gamma = log_{10}~c + \xi_1 log_{10}~A + \xi_2log_{10}~P_b$ ($\xi_1$ and $\xi_2$ are the scaling exponents) regardless of ecological contexts; however, contributing factors of increased $\gamma$ diversity - either enhanced local species richness ($\alpha$ diversity) and/or spatial variation of species composition ($\beta$ diversity) - depended on constituent species' characteristics. Our findings suggest that the dual scaling law is a pervasive yet overlooked feature of complex ecosystems with important implications for biodiversity conservation.

>>>>

Here, we propose a unified framework of ecosystem size and complexity in scaling biodiversity patterns in rivers - a prime example of complex branching ecosystems. Individual streams and rivers flow through different landscapes with distinct geological and climatic backgrounds, serving as a spatial unit of unique in-stream environments [@brownDendriticNetworkStructure2010; @petersonModellingDendriticEcological2013; @mooreEmergentStabilityLarge2015; @teruiMetapopulationStabilityBranching2018; @isaakSlowClimateVelocities2016; @mcguireNetworkAnalysisReveals2014; @abbottUnexpectedSpatialStability2018]. The recurrent merging of diverse tributaries ultimately forms a fluvial network with fractal branching patterns [@rodriguez-iturbeFractalRiverBasins2001]. As such, the complexity of branching structure, which we define here as the probability of branching per unit river distance [@teruiMetapopulationStabilityBranching2018; @yeakelSynchronisationStabilityRiver2014], may represent a macro-scale control of the ecosystem's habitat heterogeneity (habitat diversity per unit area) [@grantLivingBranchesPopulation2007; @mooreEmergentStabilityLarge2015; @teruiMetapopulationStabilityBranching2018]. Meanwhile, ecosystem size (watershed area) should determine the metacommunity size and total habitat diversity (area x heterogeneity), two factors that regulate biodiversity at the metacommunity level [@hubbellUnifiedNeutralTheory2001; @kallimanisHowDoesHabitat2008]. Hence, riverine biodiversity may manifest scaling laws along the two orthogonal dimensions of branching networks.

The present study combines theory and analysis of extensive community data from two different regions of the globe to show that multiple ecological pathways converge to the emergence of dual diversity scaling. Specifically, watershed-scale species richness ($\gamma$ diversity) increased linearly with ecosystem size $A$ and branching probability $P_b$ on logarithmic axes as $log_{10}~\gamma = log_{10}~c + \xi_1 log_{10}~A + \xi_2log_{10}~P_b$ ($\xi_1$ and $\xi_2$ are the scaling exponents) regardless of ecological contexts. However, contributing factors of increased $\gamma$ diversity - either enhanced local species richness ($\alpha$ diversity) and/or spatial variation of species composition ($\beta$ diversity) - depended on constituent species' characteristics. Our findings suggest that the dual scaling law is a pervasive yet overlooked feature of complex ecosystems with important implications for biodiversity conservation.
